A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-16 Manuscript/OldVsns/Manuscript-2.Rmd

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with smaller values signifying 'weak' phylogenetic signal, while larger values are interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical, through the use of confidence intervals [@Vandelook2019] or likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Likewise, qualitative comparisons of $\lambda$ across multiple phenotypic traits have also been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, it seems intuitive to interpret the strength of phylogenetic signal in this manner, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood ($\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under Brownian motion). However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. \hfill\break

First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

@@

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 1). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, with the worst precision at intermediate levels of signal ($\lambda_{in}\approx0.5$), and improved precision as input levels approached the extremes of its range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). Results were similar when $\lambda$ was co-estimated with regression parameters in PGLS regression (Figure 2). Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell 2010 [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal were not, and the spread of $\lambda_{est}$ was even broader than that observed when $\lambda$ was estimated for only the dependent variable. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

Adams and Collyer [-@AdamsCollyer2019] recently suggested that the strength of phylogenetic signal could be represented an effect size, obtained from $Kappa$ and its empirical sampling distribution from permutation. Here we formalize that suggestion, and find an effect size as:

@@

Similarly, an effect size based on $\lambda$ could be envisioned as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}
\end{align}

Note that under the null hypothesis, $E(\lambda)=0$, a no phylogenetic signal is expected under this condition [@Freckleton_et_al2002]. \hfill\break

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 3. Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than to estimates of $Z_\lambda$. Second, the precision of $Z_K$ is considerably more stable, as coefficients of variation for the set of $\sigma^2_{Z_K}$ across input levels of phylogenetic signal for the same sample size was an order of magnitude smaller than that observed for $\sigma^2_{Z_\lambda}$ (Figure 3). This implies that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

[insert Figure 3 here]

>>>>

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with smaller values signifying 'weak' phylogenetic signal, while larger values are interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical. For instance, some have evaluated whether the observed $\lambda$ differs from some expected value through the use of confidence intervals [@Vandelook2019] or by performing likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Additionally, qualitative comparisons of $\lambda$ estimates obtained from multiple phenotypic traits have been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, statements regarding the strength of phylogenetic signal based on $\lambda$ are rather common in the evolutionary literature. We conducted a literature survey in Google.scholar and found that of the 204 papers published in 2019 that estimated and reported Pagel's $\lambda$, 40% interpreted the strength of phylogenetic signal for at least one phenotypic trait. Additionally, nearly 30% of the 421 $\lambda$ estimates between 0.25 and 0.75 were assigned a strength of signal, and because nearly half of the 1,572 values reported were near the limits of the parameter (Figure 1), this percentage is even higher, as biological interpretation of phylogenetic signal at the limits of $\lambda$ are known. \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal based on the value of $\lambda$, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood. Specifically, $\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under Brownian motion. However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., the number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also assumes that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we still lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

@@

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, with the worst precision at intermediate levels of signal ($\lambda_{in}\approx0.5$), and improved precision as input levels approached the extremes of its range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). Results were similar when $\lambda$ was co-estimated with regression parameters in PGLS regression (Figure 3). Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell 2010 [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal were not, and the spread of $\lambda_{est}$ was even broader than that observed when $\lambda$ was estimated for only the dependent variable. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

Adams and Collyer [-@AdamsCollyer2019] recently suggested that the strength of phylogenetic signal could be represented as an effect size, based on the $Kappa$ statistic and its empirical sampling distribution from permutation. Here we formalize that suggestion, and find an effect size as:

@@

where $K_{obs}$ is the observed phylogenetic signal, and $\hat\mu_K$ and $\hat\sigma_K$ are the mean and standard deviation of the empirical sampling distribution of $Kappa$ obtained via permutation. Similarly, an effect size based on $\lambda$ could be envisioned as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}.
\end{align}

In this case, $\lambda_{obs}$ and $\hat\sigma_\lambda$ are empirically derived using maximum likelihood. Note also that under the null hypothesis, $E(\lambda)=0$, a no phylogenetic signal is expected under this condition [@Freckleton_et_al2002]. \hfill\break

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 4. Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than do estimates of $Z_\lambda$. Second, the precision of $Z_K$ is considerably more stable as compared with $Z_\lambda$, as coefficients of variation for the set of $\sigma^2_{Z_K}$ across input levels of phylogenetic signal were an order of magnitude smaller for than was observed for $\sigma^2_{Z_\lambda}$ (Figure 4). This implied that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

[insert Figure 4 here]
